Chain separation distribution in protein contact networks

### 2022-04-14 paper/main.tex
Updates from Overleaf

\begin{document}
\title{Chain separation distribution in protein contact networks}
\title{Theoretical approximation of chain separation distribution in protein contact networks}
\title{Theory of chain separation distributions gives insights in real protein contact networks}
%\title{}
\author{Nora Molkenthin}
\affiliation{Potsdam Institut für Klimafolgenforschung}
\affiliation{Center for Advancing Electronics Dresden (cfaed), Technical University of Dresden}
\author{J. J.  Güven}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}
\author{Steffen Mühle}
\affiliation{Max-Planck Institute for Dynamics and Self-Organization, 37077 Göttingen, Germany}
\author{Antonia S J S Mey}
\email[Electronic Address: ]{antonia.mey@ed.ac.uk}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}

\begin{abstract}
The mechanisms by which we can determine a protein's 3D structure based on the protein's amino acid sequence have long been one of the key mysteries of biophysics. With the advent of AlphaFold2, making predictions for protein structures based on amino acid sequences at an accuracy similar to experiments is now possible, yet requires a lot of available structural data for training the model. Predicting contact maps from sequence \textit{de novo} is an appealing alternative. Here, we are interested in understanding the emergent behavior of protein contact maps derived from a network ensemble perspective of a geometrically constrained random interaction model. We show that the frequency distributions of amino acid distances obtained from simulations of the geometrically constrained interaction model can be described by an analytical approximation of the two-dimensional analogue of this model. To this end, we look at the scaling behavior of the simulation with the analytical approximation for different chain lengths typically found in proteins with good agreement. Furthermore, we compare our constrained model with frequency distributions of amino acid distances from contact maps of real proteins---from the protein data bank and AlphaFold2 database---showing that these follow a similar trend as the geometrically constrained model.   
 We therefore suggest the conclusion that the approximate power-law behavior of the amino acid distance distribution can be attributed to geometric constraints of protein chains and amino acid sequences only play a secondary role.  
%The distribution of chain separations of interacting amino acids in proteins roughly follows a power law. Here we show, analytically and in simulations, that a geometrical stochastic model of folding chains can explain this behavior. 
\end{abstract}
\maketitle

\section{Introduction}
Proteins, the molecular machines of every living organism perform vital tasks required for life to persist, ranging from transport (e.g. hemoglobin)~\cite{ahmed2020hemoglobin}, signal transduction (e.g. rhodopsin)~\cite{nagata2021rhodopsins}, immune responses (e.g. antibodies), and hormonal regulation (e.g. insulin)~\cite{dill2008protein,Dill1042}. All natural proteins are made of 20 different amino acids which dictate the three-dimensional conformations the proteins adopt in order to function~\cite{scheraga2007protein}. One of the great challenges has been to understand how the primary structure, the amino acid sequence, can lead to the fully folded functional protein. The last 50 years has seen many different routes to computationally predict biologically active protein structures without having to solve a crystal structure of the protein. [input information from Rohan's presentation]. AlphaFold2, a machine learning model for structure prediction, has shown predictiveness at a level of experimental accuracy in the most recent protein structure prediction challenge~\cite{jumper2021highly, kryshtafovych2021critical}. This has lead to an explosion of available structural protein models beyond the protein database, containing structures from crystallographic, NMR, and cryo-EM experiments~\cite{berman2000protein}. Beyond the prediction of structure understanding emergent patterns in proteins on a more fundamental level is of interest, as it opens up a deeper understanding of proteins structure and function relationship.  One way of looking at the functional forms of proteins and classifying their structural properties is by looking at protein contact maps (PCM) or protein contact networks (PCN)~\cite{Vendruscolo2002,DiPaola2013,Estrada2011}. A PCM---we will use PCMs in favor of PCN in this paper---is the resulting adjacency matrix of spatially close amino acids, given a certain cut-off as determined typically from a protein structure. PCMs have shown promise in understanding protein folding patterns as well as revealing allosteric communication pathways~\cite{yao2019establishing, menichetti2016network,dokholyan2002topological}. Their use is also prominently featured in machine learning approaches. One challenge is to predict a PCM \textit{de novo} from sequence alone~\cite{bassot2019using, rives2021biological, rao2021msa}, without using structural data as part of the training set, as was done for AlphaFold2. These PCMs can then be used in e.g. graph convolutional machine learning approaches to train a model able to predict binding affinities for certain types of proteins~\cite{jiang2020drug},
From a physicist's perspective it is interesting to understand the emergent behavior in protein structure by looking at PCMs. Since a PCM is an adjacency matrix and therefore a graph, it is possible to investigate PCMs using common network measures. For example, Bartoli \textit{et al.}~\cite{bartoli2008effecta} build a model in which they assume that proteins have an amino acid distance distribution that follows $\frac{1}{s}$, where $s$ is the separation of two connected amino acids along the backbone chain (in the following called \emph{amino acid distance}) as the simplest implementation of a connection probability, thus decreasing with the distance along the sequence. However, they do not provide a quantitative explanation for this assumption.

@@

Based on such considerations, the network formation process repeatedly connects random pairs of units, without causing any intersections along the chain~\cite{molkenthin2016scaling}. Such a model can now be used to investigate the emergent behavior of amino acid distance distributions as used heuristically by Bartoli \textit{et al.}~\cite{bartoli2008effecta}. To this end, we define and analyze the amino acid distance distributions $P(s)$ that emerge from the geometric model, where the \emph{amino acid distance} $s$ of a link is defined as the distance of the two end units along the original chain.

This distribution is then compared to the amino acid distance distribution extracted from protein contact networks generated from protein data bank data.

We find that, despite the highly simplified model, the generated amino acid distance distribution is in agreement with the one observed in networks generated from PDB data and roughly consistent with the heuristic $1/s$ finding. The analytical approximation provides a correction term on top of the $1/s$ heuristic.

%%%%%%%%%%%%%%%%%
% Section       %
%%%%%%%%%%%%%%%%%
\section{Protein contact maps}
The complex interaction pattern between the amino acids in a protein can be very naturally expressed as a network or graph, in which each amino acid is represented by a node and spatial proximity is encoded as a link. Whenever two central $C_\alpha$ atoms are closer together than a threshold $d_c$, they are connected, linked, or in `close contact'. These connections can be determined from the 3D functional or folded structure of the protein and are encoded in a so called contact map. This contact map is effectively an adjacency matrix $\textbf{A}^{\mathrm{struc}}$ or binary matrix
\begin{equation}
  A^{\mathrm{struc}}_{ij}=
  \begin{cases}
   0, & \text{ if } d_{i,j}>d_c \text{ or } i=j\\
      1, & \text{ if } d_{i,j}\leq d_c .
      \end{cases}
    \label{eq:aij}
\end{equation}
Adjacency matrices are commonly used to describe networks and make up a \textit{protein contact map}. 
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/Fig1/Fig1.pdf}
    \caption{a) The amino acid distance $s$ is defined as the separation along the chain (teal) of two connected units (pink). b)~The solid green link prohibits some connections from node~$A$ (dotted pink), while others are still permitted (dotted green). Counting all newly excluded links, we find that a link of length $s$ prohibits $2(l-1)$ links of length $l$ up to a maximum of $s$ and $2(s-1)$ links of length $s$ or more. c)~Example of a protein contact map taken from a 2D simulation. Pink squares indicate a contact is present between two amino acids.}
    \label{fig:schematic}
\end{figure}

@@

The underlying formation mechanisms of individual protein folds are incredibly complex and depend on the surrounding solvent, as well as the specific amino acid sequence and their quantum-mechanical interactions. Here we want to assess if the geometrically constrained protein model's contact map distributions capture real protein behavior as suggested in~\cite{bartoli2008effecta}. To understand this better, we want to study the ensemble of PCMs from real protein structures. To this end we will look at `averaged' contact maps over many different proteins that have different amino acid sequences, but their overall sequence or chain length is the same. 
We will use the RCSB protein data bank (PDB), a database of structurally resolved protein amino acid sequences through X-ray crystallography, NMR or cryo-EM experiments and the synthetically generated structural data base from the machine learned structures of AlphaFold2~\cite{jumper2021highly}. The PDB contains just under 200,000 protein structures to date~\cite{PDB}. The distribution of amino acid chain lengths in the PDB is illustrated in Fig.~\ref{fig:pdb_stats} b. Protein chain lengths in the PDB vary from small fragments of less than 10 amino acids to large agglomerates of over 2000 amino acids. The most frequently occurring chain lengths, however, are between 100 and 300 amino acids as shown in Fig.~\ref{fig:pdb_stats} b. The pink subset of the barplot shown in Fig.~\ref{fig:pdb_stats} b, are the set of structures used that fulfill the criterion on amino acid chain lengths between 85 and 115 amino acids, 185 and 215 amino acids and 285 and 315 amino acids and represent all structures for the analysis. The second database consists of synthetically generated structures using AlphaFold2 with sequences taken from SwissProt~\cite{alpha fold 2 database}. SwissProt is a manually curated database of protein sequences containing over 500,000 protein sequences~\cite{SwissProt}, whose amino acid chain length distribution is shown in Fig.~\ref{fig:pdb_stats} a in blue-green. The pink bars again indicate the subset of AlphaFold2 structures used for the contact map analysis.

 \begin{figure}[t]
        \centering
	\includegraphics[width=0.8\columnwidth]{paper/figures/Fig2/Fig2.pdf}
	    \caption{Histogram of protein sequence lengths for protein structures from two structure databases. a) AlphaFold2 predicted protein structures ($\alpha$F) taken from all protein sequences in the reviewed Uniprot Knowledge Base (Swiss-Prot) database based on their Uniprot ids. Pink bars shows fraction of structures used for the average contact map analysis.  b) RSCB PDB chain length frequencies based on sequence length. Pink bars in plot show proteins structures whose sequence lengths matches the resolution of structure accounting for the whole sequence. Only pink PDB samples were used for the average contact map analysis. c) Example of a protein contact map taken from PDB ID: 1AKI, with a cut-off of 8~Å. Pink squares indicate a contact present between amino acids.}
        \label{fig:pdb_stats}
\end{figure}
All structure based contact maps are generated in the same way for AlphaFold2 structures and PDB structures. Structures used were downloaded in pdb format, read in with MDAnalysis version 2.0.0~\cite{gowers2016mdanalysis} and only C$_{\alpha}$ were selected and used for the analysis. Proteins were categorised by length and placed into three groups. Amino acid chain lengths between 85 and 115 amino acids, 185-215 amino acids and 285-315. For each of the used sequences, all pairwise C$_{\alpha}$ distances were assessed and two C$_{\alpha}$ atoms were in contact if their distance was less than 8~Å. All pairwise distances were evaluated and in this way, building up the binary contact map or adjacency $A_{ij}^{\mathrm{struc}}$.  
All code for the construction of the contact maps both for the geometrically constrained system and structural protein data and their analysis can be found on github at: \url{https://github.com/meyresearch/sequence_distance_distribution}.

@@

Let us start by introducing the auxiliary variable $F_k(s)$, defined to be the number of possible links with an amino acid distance of $s$ if $k$ links have been added before. Before any links are added, all amino acid distances are equally likely, as there are $F_0(s)=N$ possibilities of making a link of each distance $s$ with $2\leq s < N/2$.

@@

Together with the initial condition $F_0(s)=N$, this expression constitutes an iteration rule from which $F_k(s)$ can be found for all $k$ and $s$. Aiming to decouple those dynamics for different $s$, we make a heuristic guess for $\Pin^k(s_k)$. It consists of two separate approximation which are used in steps $k<a$ and steps $k\geq a$ respectively, where the threshold $1\leq a\leq N/2$ is a free parameter. In the early stage we use a uniform probability
\begin{equation}
    \Pin^{k\ll N/2}(s)\approx \Pin^{k=0}(s)=\frac{2}{N}.
    \label{eq.P_small}
\end{equation}
For later steps, the probability of a link to still be in the pool decreases with $s$. We thus approximate $\Pin(s)$ with an ansatz for the $k-$average, namely, that it drops off as $\frac{1}{s}$, leading to
\begin{equation}
    \Pin^{k\gg 1}(s)\approx \sum_{k=a}^{N-4}\Pin^{k}(s)
    \approx \frac{s^{-1}}{ (H_{\frac{N}{2}}-1)},
    \label{eq.P_large}
\end{equation}
where $H_{\frac{N}{2}}$ is the harmonic number serving as a normalization factor. 
As we will see later this approximation holds most precisely for intermediate values of $k$, with the $k-$average of $\Pin^{k\gg 1}(s)\approx P(s)$, making it self-consistent. Errors for smaller and larger values of $k$ are thought to approximately cancel out.
Substituting Eq.~\ref{eq.P_small} and Eq.~\ref{eq.P_large} into Eq.~\ref{eq.Fk_rec} results in the following two recursions for the early and late evolution of the amino acid distance distribution respectively.
\begin{align}
   F_{k\ll N/2}(s)&\approx 
   F_{k-1}(s)(1-\frac{2}{N} (s-1) 
   +\frac{4}{N^2} s (s-1)\nonumber \\
   &-\frac{2}{N^2}(s^2+s-2)\nonumber \\
   &=F_{k-1}(s)(1-\flow(s)),
   \label{eq.Fk_rec_low}
\end{align}
where 
\[\flow(s)=-\frac{2}{N^2}s^2+\left(\frac{2}{N}+\frac{6}{N^2}\right)s-\left(\frac{2}{N}+\frac{4}{N^2}\right),\]
and

@@

In 2D this is repeated for $N-3$ steps until no more links can be added \cite{molkenthin2016scaling}.
The probability distribution of the realized amino acid distances of all added links is then given by the average over the available pools at each link addition step, leading to
\begin{align}
    P(s)&=\frac{1}{N-3}\sum_{k=0}^{N-4} \frac{F_k(s)}{C_k} \nonumber \\
    &\approx \frac{N}{N-3}\left(\frac{1-\flow(s)}{1-\fhigh(s)}\right)^a
\sum_{k=0}^{N-4} \frac{1}{\evdel{C_k}_k} \left(1-\fhigh(s)\right)^{k} \nonumber \\
    &\approx \left(\frac{1-\flow(s)}{1-\fhigh(s)}\right)^a\frac{A}{\fhigh(s)},
    \label{eq.Ps}
\end{align}
where $\evdel{C_k}_k$ is the average of $C_k$ over $k$.
This approximation allows the use of the geometric series for solving the equation. 
The resulting expression in Eq.~\ref{eq.Ps} is close to a power-law in $s$ with an exponent of \red{something close to -1.2 depending on a}, thus justifying the earlier approximation of $\Pin^k(s)$.
%which makes use of Eq.~\ref{eq.Pk}. 
%The only unknown in this expression is $C_k$, \red{which we can not compute exactly...?}. However, since $C_k$ monotonically decreases with $k$, $N>C_k>C_{N-2}$ always holds and we can obtain an upper and lower bound by replacing all $C_k$ in the expression with either $N$ (underestimates all probabilities) or $C_{N-2}$ (overestimates all probabilities). Each time we get an expression of the form 
% \begin{align}
%      P(s)&\approx\frac{N}{N-3}\sum_{k=0}^{N-3} \frac{1}{C_k}\prod_{i}^k\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)\nonumber \\
%      &=\frac{N}{N-3} \sum_{k=0}^{N-3}\frac{1}{C_k}\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)^k \nonumber \\
%      &\approx \frac{A}{f(s)}
% \end{align}
% for $1<s<\frac{N}{2}$,
% where $A$ is an unknown constant, containing an approximation of the various normalization constants $C_i$, as well as the direct effects of the chain length.

@@

\section{Results}
\subsection{Analytic approximation and 2D simulations}
Fig.~\ref{fig:2d_sim} compares Eq.~\ref{eq.Ps} with respect to the two-dimensional simulation data introduced in~\cite{molkenthin2020self}. The simulation data is generated from 10 repeats with chain lengths of up to 498 amino acids long. The simulation data is plotted showing the mean frequencies of each amino acid distance up to $N/2\sim250$. The shaded area represent the 95\% confidence interval calcualted from the 10 repeats. The theoretical approximation from Eq.~\ref{eq.Ps} was fitted to the simulation data. The two fitting parameters were chosen to be $a=6$ and $A=0.39$, to best capture the trend of the simulation data. The fitted approximation captures the behavior of the simulation data well, both describe two dimensional systems. 
 \begin{figure}[tb]
        \centering
	\includegraphics[width=0.45\textwidth]{paper/figures/Fig3/Fig3.pdf}
	    \caption{Amino acid distance distribution for two-dimensional simulated data plotted against the analytic approximation. Green area shows the 95\% confidence interval.}
        \label{fig:2d_sim}
\end{figure}

\subsection{3D simulations and PCMs from structural data }
Next, we investigate if and how the analytical approximation can also capture trends in 3-D simulation data, as well as experimentally determined protein structures. This seeks to answer the question if indeed a power law can be used to describe the probability distribution of amino acid distances. Fig.~\ref{fig:sdd} summarises the results for protein chain lengths of $\approx$100 (a), $\approx$200, and $\approx$ 300 amino acids, protein chain lengths, most commonly found in living organism (see. Fig.~\ref{fig:pdb_stats} a). We are comparing data from repeats of 30 simulation runs of the three-dimensional simulations, as introduced in~\cite{molkenthin2020self}, and protein contact maps from RCSB~\cite{}, and AlphaFold2~\cite{}.

>>>>

\begin{document}
\title{Chain separation distribution in protein contact networks}
\title{Theoretical approximation of chain separation distribution in protein contact networks}
\title{Theory of chain separation distributions gives insights in real protein contact networks}
%\title{}
\author{Nora Molkenthin}
\affiliation{Potsdam Institut für Klimafolgenforschung}
\affiliation{Center for Advancing Electronics Dresden (cfaed), Technical University of Dresden}
\author{J. J.  Güven}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}
\author{Steffen Mühle}
\affiliation{Max-Planck Institute for Dynamics and Self-Organization, 37077 Göttingen, Germany}
\author{Antonia S J S Mey}
\email[Author to whom correspondence should be addressed:]{antonia.mey@ed.ac.uk}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}

\begin{abstract}
The mechanisms by which we can determine a protein's 3D structure based on its amino acid sequence have long been one of the key mysteries of biophysics. With the advent of AlphaFold~2, making predictions for protein structures based on amino acid sequences at an accuracy similar to experiments is now possible, yet requires a lot of available structural data for training the model. Predicting contact maps from sequence \textit{de novo} is an appealing alternative. Here, we are interested in understanding the emergent behavior of protein contact maps derived from a network ensemble perspective of a geometrically constrained random interaction model. We show that the frequency distributions of amino acid distances obtained from simulations of the geometrically constrained interaction model can be described by an analytical approximation of the two-dimensional analogue of this model. To this end, we look at the scaling behavior of the simulation with the analytical approximation for different chain lengths typically found in proteins with good agreement. Furthermore, we compare our constrained model with frequency distributions of amino acid distances from contact maps of real proteins---from the protein data bank and AlphaFold~2 database---showing that these follow a similar trend as the geometrically constrained model.   
 We therefore suggest the conclusion that the approximate power-law behavior of the amino acid distance distribution can be attributed to geometric constraints of protein chains and amino acid sequences only play a secondary role.  
%The distribution of chain separations of interacting amino acids in proteins roughly follows a power law. Here we show, analytically and in simulations, that a geometrical stochastic model of folding chains can explain this behavior. 
\end{abstract}
\maketitle

\section{Introduction}
Proteins, the molecular machines of every living organism perform vital tasks required for life to persist, ranging from transport (e.g. hemoglobin)~\cite{ahmed2020hemoglobin}, signal transduction (e.g. rhodopsin)~\cite{nagata2021rhodopsins}, immune responses (e.g. antibodies), and hormonal regulation (e.g. insulin)~\cite{dill2008protein,Dill1042}. All natural proteins are made of 20 different amino acids which dictate the three-dimensional conformations the proteins adopt in order to function~\cite{scheraga2007proteinfolding}. One of the great challenges has been to understand how the primary structure, the amino acid sequence, can lead to the fully folded functional protein. The last 50 years has seen many different routes to computationally predict biologically active protein structures without having to solve a crystal structure of the protein. [input information from Rohan's presentation]. AlphaFold~2, a machine learning model for structure prediction, has shown predictiveness at a level of experimental accuracy in the most recent protein structure prediction challenge~\cite{jumper2021highly, kryshtafovych2021critical}. This has led to an explosion of available structural protein models beyond the protein database, containing structures from crystallographic, NMR, and cryo-EM experiments~\cite{berman2000protein}. Beyond the prediction of structure understanding emergent patterns in proteins on a more fundamental level is of interest, as it opens up a deeper understanding of proteins structure and function relationship.  One way of looking at the functional forms of proteins and classifying their structural properties is by looking at protein contact maps (PCM) or protein contact networks (PCN)~\cite{Vendruscolo2002,dipaola2013protein,Estrada2011}. A PCM---we will use PCMs in favor of PCN in this paper---is the resulting adjacency matrix of spatially close amino acids, given a certain cut-off as determined typically from a protein structure. PCMs have shown promise in understanding protein folding patterns as well as revealing allosteric communication pathways~\cite{yao2019establishing, menichetti2016network,dokholyan2002topological}. Their use is also prominently featured in machine learning approaches. One challenge is to predict a PCM \textit{de novo} from sequence alone~\cite{bassot2019using, rives2021biological, rao2021msa}, without using structural data as part of the training set, as was done for AlphaFold~2. These PCMs can then be used in e.g. graph convolutional machine learning approaches to train a model able to predict binding affinities for certain types of proteins~\cite{jiang2020drug},
From a physicist's perspective it is interesting to understand the emergent behavior in protein structure by looking at PCMs. Since a PCM is an adjacency matrix and therefore a graph, it is possible to investigate PCMs using common network measures. For example, Bartoli \textit{et al.}~\cite{bartoli2008effecta} build a model in which they assume that proteins have an amino acid distance distribution that follows $s^{-1}$, where $s$ is the separation of two connected amino acids along the backbone chain (in the following called \emph{amino acid distance}) as the simplest implementation of a connection probability, thus decreasing with the distance along the sequence. However, they do not provide a quantitative explanation for this assumption.

@@

Based on such considerations, the network formation process repeatedly connects random pairs of units, without causing any intersections along the chain~\cite{molkenthin2016scaling}. Such a model can now be used to investigate the emergent behavior of amino acid distance distributions as used heuristically by Bartoli \textit{et al.}~\cite{bartoli2008effecta}. To this end, we define and analyze the amino acid distance distributions $P(s)$ that emerge from the geometric model, where the \emph{amino acid distance} $s$ of a link is defined as the distance of the two end units along the original chain. This distribution is then compared to the amino acid distance distribution extracted from protein contact networks generated from protein data bank and AlphaFold~2 data.

We find that, despite the highly simplified model, the generated amino acid distance distribution is in agreement with the one observed in networks generated from PDB data and roughly consistent with the heuristic $s^{-1}$ finding. The analytical approximation provides a correction term on top of the $s^{-1}$ heuristic.

%%%%%%%%%%%%%%%%%
% Section       %
%%%%%%%%%%%%%%%%%
\section{Protein contact maps}
\label{sec:methods}
The complex interaction pattern between the amino acids in a protein can be very naturally expressed as a network or graph, in which each amino acid is represented by a node and spatial proximity is encoded as a link. Whenever two central $C_\alpha$ atoms are closer together than a threshold $d_c$, they are connected, linked, or in `close contact'. These connections can be determined from the 3D functional or folded structure of the protein and are encoded in a so-called contact map. This contact map is effectively an adjacency matrix $\textbf{A}^{\mathrm{struc}}$ or binary matrix
\begin{equation}
  A^{\mathrm{struc}}_{ij}=
  \begin{cases}
   0, & \text{ if } d_{i,j}>d_c \text{ or } i=j\\
      1, & \text{ if } d_{i,j}\leq d_c.
      \end{cases}
    \label{eq:aij}
\end{equation}
Adjacency matrices are commonly used to describe networks and make up a \textit{protein contact map}. 
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/Fig1/Fig1.pdf}
    \caption{a) The amino acid distance $s$ is defined as the separation along the chain (teal) of two connected units (pink). b)~The solid green link prohibits some connections from node~$A$ (dotted pink), while others are still permitted (dotted green). Counting all newly excluded links, we find that a link of length $s$ prohibits $2(l-1)$ links of length $l$ up to a maximum of $s$ and $2(s-1)$ links of length $s$ or more.}
    \label{fig:schematic}
\end{figure}

@@

The underlying formation mechanisms of individual protein folds are incredibly complex and depend on the surrounding solvent, as well as the specific amino acid sequence and their quantum-mechanical interactions. Here we want to assess if the geometrically constrained protein model's contact map distributions capture real protein behavior as suggested in~\cite{bartoli2008effecta}. To understand this better, we want to study the ensemble of PCMs from real protein structures. To this end we will look at `averaged' contact maps over many different proteins that have different amino acid sequences, but their overall sequence or chain length is the same. 
We will use the RCSB protein data bank (PDB), a database of structurally resolved protein amino acid sequences through X-ray crystallography, NMR or cryo-EM experiments and the synthetically generated structural database from the machine learned structures of AlphaFold~2~\cite{jumper2021highly}. The PDB contains just under 200,000 protein structures to date~\cite{PDB}. The distribution of amino acid chain lengths in the PDB is illustrated in Fig.~\ref{fig:pdb_stats} b. Protein chain lengths in the PDB vary from small fragments of less than 10 amino acids to large agglomerates of over 2000 amino acids. The most frequently occurring chain lengths, however, are between 100 and 300 amino acids as shown in Fig.~\ref{fig:pdb_stats} b. The pink subset of the barplot shown in Fig.~\ref{fig:pdb_stats} b, are the set of structures used that fulfill the criterion on amino acid chain lengths between 85 and 115 amino acids, 185 and 215 amino acids and 285 and 315 amino acids and represent all structures for the analysis. The second database consists of synthetically generated structures using AlphaFold~2 with sequences taken from SwissProt~\cite{alpha fold 2 database}. SwissProt is a manually curated database of protein sequences containing over 500,000 protein sequences~\cite{SwissProt}, whose amino acid chain length distribution is shown in Fig.~\ref{fig:pdb_stats} a in teal. The pink bars again indicate the subset of AlphaFold~2 structures used for the contact map analysis. In this case, the used structures were filtered first with the condition that the AlphaFold~2 pre-residue confidence score~\cite{jumper2021highly} was `Very high', i.e. above 90 out of 100, and then by removing any structures describing the same protein to remove duplicates.

 \begin{figure}[t]
        \centering
	\includegraphics[width=0.8\columnwidth]{paper/figures/Fig2/Fig2.pdf}
	    \caption{Histogram of protein sequence lengths for protein structures from two structure databases. a) AlphaFold~2~\cite{jumper2021highly} predicted protein structures taken from protein sequences in the reviewed Uniprot Knowledge Base (Swiss-Prot) database based on their Uniprot IDs. Pink bars show fraction of structures used for the average contact map analysis.  b) RSCB~\cite{PDB} PDB chain length frequencies based on sequence length. Pink bars in plot show protein structures whose sequence lengths matches the resolution of structure accounting for the whole sequence. Only pink PDB samples were used for the average contact map analysis. c) Example of a protein contact map taken from PDB ID: 3UFG, with a cut-off of 8~Å. Pink squares indicate a contact present between amino acids.}
        \label{fig:pdb_stats}
\end{figure}
All structure-based contact maps are generated in the same way for AlphaFold~2 structures and PDB structures. Structures used were downloaded in pdb format, read in with MDAnalysis version 2.0.0~\cite{gowers2016mdanalysis} and only C$_{\alpha}$ were selected and used for the analysis. Proteins were categorised by length and placed into three groups. Amino acid chain lengths between 85 and 115 amino acids, 185-215 amino acids and 285-315. For each of the used sequences, all pairwise C$_{\alpha}$ distances were assessed and two C$_{\alpha}$ atoms were in contact if their distance was less than 8~Å. All pairwise distances were evaluated and in this way, building up the binary contact map or adjacency $A_{ij}^{\mathrm{struc}}$.  
All code for the construction of the contact maps both for the geometrically constrained system and structural protein data and their analysis can be found on GitHub at: \url{https://github.com/meyresearch/sequence_distance_distribution}.

@@

Let us start by introducing the auxiliary variable $F_k(s)$, defined to be the number of possible links with an amino acid distance of $s$ if $k$ links have been added before. Before any links are added, all amino acid distances are equally likely, as there are $F_0(s)=N$ possibilities of making a link of each distance $s$ with $2\leq s < \frac{N}{2}$.

@@

Together with the initial condition $F_0(s)=N$, this expression constitutes an iteration rule from which $F_k(s)$ can be found for all $k$ and $s$. Aiming to decouple those dynamics for different $s$, we make a heuristic guess for $\Pin^k(s_k)$. It consists of two separate approximations which are used in steps $k<a$ and steps $k\geq a$ respectively, where the threshold $1\leq a\leq N/2$ is a free parameter. In the early stage we use a uniform probability
\begin{equation}
    \Pin^{k\ll N/2}(s)\approx \Pin^{k=0}(s)=\frac{2}{N}.
    \label{eq.P_small}
\end{equation}
For later steps, the probability of a link to still be in the pool decreases with $s$. We thus approximate $\Pin(s)$ with an ansatz for the $k-$average, namely, that it drops off as $\frac{1}{s}$, leading to
\begin{equation}
    \Pin^{k\gg 1}(s)\approx \sum_{k=a}^{N-4}\Pin^{k}(s)
    \approx \frac{s^{-1}}{ (H_{\frac{N}{2}}-1)},
    \label{eq.P_large}
\end{equation}
where $H_{\frac{N}{2}}$ is the harmonic number serving as a normalization factor. 
As we will see later this approximation holds most precisely for intermediate values of $k$, with the $k-$average of $\Pin^{k\gg 1}(s)\approx P(s)$, making it self-consistent. Errors for smaller and larger values of $k$ are thought to approximately cancel out.
Substituting Eq.~\ref{eq.P_small} and Eq.~\ref{eq.P_large} into Eq.~\ref{eq.Fk_rec} results in the following two recursions for the early and late evolution of the amino acid distance distribution respectively.
\begin{align}
   F_{k\ll N/2}(s)&\approx 
   F_{k-1}(s)(1-\frac{2}{N} (s-1) 
   +\frac{4}{N^2} s (s-1)\nonumber \\
   &-\frac{2}{N^2}(s^2+s-2)\nonumber \\
   &=F_{k-1}(s)(1-\flow(s)),
   \label{eq.Fk_rec_low}
\end{align}
where 
\[\flow(s)=-\frac{2}{N^2}s^2+\left(\frac{2}{N}+\frac{6}{N^2}\right)s-\left(\frac{2}{N}+\frac{4}{N^2}\right),\]
and

@@

In 2D this is repeated for $N-3$ steps until no more links can be added \cite{molkenthin2016scaling}.
The probability distribution of the realized amino acid distances of all added links is then given by the average over the available pools at each link addition step, leading to
\begin{align}
    P(s)&=\frac{1}{N-3}\sum_{k=0}^{N-4} \frac{F_k(s)}{C_k} \nonumber \\
    &\approx \frac{N}{N-3}\left(\frac{1-\flow(s)}{1-\fhigh(s)}\right)^a
\sum_{k=0}^{N-4} \frac{1}{\evdel{C_k}_k} \left(1-\fhigh(s)\right)^{k} \nonumber \\
    &\approx \left(\frac{1-\flow(s)}{1-\fhigh(s)}\right)^a\frac{A}{\fhigh(s)},
    \label{eq.Ps}
\end{align}
where $\evdel{C_k}_k$ is the average of $C_k$ over $k$.
This approximation allows the use of the geometric series for solving the equation. 
The resulting expression in Eq.~\ref{eq.Ps} is by visual inspection of Fig.\ref{fig:2d_sim} resembles a power law with an exponent of -1, thus justifying the earlier approximation of $\Pin^k(s)$.
%which makes use of Eq.~\ref{eq.Pk}. 
%The only unknown in this expression is $C_k$, \red{which we can not compute exactly...?}. However, since $C_k$ monotonically decreases with $k$, $N>C_k>C_{N-2}$ always holds and we can obtain an upper and lower bound by replacing all $C_k$ in the expression with either $N$ (underestimates all probabilities) or $C_{N-2}$ (overestimates all probabilities). Each time we get an expression of the form 
% \begin{align}
%      P(s)&\approx\frac{N}{N-3}\sum_{k=0}^{N-3} \frac{1}{C_k}\prod_{i}^k\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)\nonumber \\
%      &=\frac{N}{N-3} \sum_{k=0}^{N-3}\frac{1}{C_k}\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)^k \nonumber \\
%      &\approx \frac{A}{f(s)}
% \end{align}
% for $1<s<\frac{N}{2}$,
% where $A$ is an unknown constant, containing an approximation of the various normalization constants $C_i$, as well as the direct effects of the chain length.

@@

\section{Results}
\subsection{Analytic approximation and 2D simulations}
Fig.~\ref{fig:2d_sim} compares Eq.~\ref{eq.Ps} with respect to the two-dimensional (a) and three-dimensional (b) simulation data introduced in~\cite{molkenthin2020self}. The 2D simulation data, generated from 10 repeats with chain lengths of up to 498 amino acids, consists of frequencies of amino acid distances in each simulation run, while the 3D simulation data consists of adjacency matrices computed from 30 simulation runs, and was analyzed the same way as the PDB and AlphaFold~2 protein data. These are plotted showing the mean frequencies of each amino acid distance up to $\frac{N}{2}\sim250$ and $\frac{N}{2}\sim150$ for 2D and 3D simulations, respectively. The shaded areas represent the 95\% confidence intervals. The theoretical approximation from Eq.~\ref{eq.Ps} was fitted to both the 2D and 3D simulation data separately. The two parameters, the exponent $a$, and the dimensionality constant $A$, were given by the fit as $a_{\mathrm{2D}}=6 \pm 1$, $A_{\mathrm{2D}}=(3.0 \pm 0.5)\cdot 10^{-4}$ and $a_{\mathrm{3D}}=1 \pm 1$, $A_{\mathrm{3D}}=(7 \pm 2)\cdot 10^{-3}$ for 2D and 3D simulations, respectively. The fitted approximation captures the behavior of the simulation data well, as in the case for 2D simulations, both describe two-dimensional systems.

 \begin{figure}[htb]
        \centering
	\includegraphics[width=\columnwidth]{paper/figures/Fig3/Fig3.pdf}
	    \caption{Amino acid distance distributions for a) two-dimensional and b) three-dimensional simulated data plotted against the analytic approximation. Pink area shows the 95\% confidence interval. c) Example of a protein contact map taken from one of the three-dimensional simulation runs. The pink squares indicate a contact between two nodes.}
        \label{fig:2d_sim}
\end{figure}
